---
title: "Workshop Contents"
output: html_document
---

This **pointblank** workshop will teach you *a lot* about what **pointblank** can do, and, it'll give you an opportunity to experiment with the package.

The goal of the workshop is to introduce you to a lot of examples and provide some time to use the functions of **pointblank** with some sample datasets, learning bit-by-bit as we go.

Each module of the workshop focuses on a different subset of functions, with relevant examples and exercises in each. They are all presented here as **R Markdown** (.Rmd) files with one file for each workshop module:

- `"01-intro-to-data-validation.Rmd"` (The `agent`, validation fns, interrogation/reports)

- `"01z-PRACTICE-01.Rmd"` (Practice with functions from module `01`)

- `"02-scan-your-data.Rmd"` (Looking at your data with `scan_data()`)

- `"03-expect-test-functions.Rmd"` (Using the `expect_*()` and `test_*()` functions)

- `"04-scaling-up-data-validation.Rmd"` (The `multiagent` and its reporting structures)

- `"04z-PRACTICE-02-03-04.Rmd"` (Practice with functions from modules `02`, `03`, and `04`)

- `"05-intro-to-data-documentation.Rmd"` (The `informant` and describing your data)

- `"06-getting-deeper-into-data-documenting.Rmd"` (Using snippets and text tricks)

- `"06z-PRACTICE-05-06.Rmd"` (Practice with functions from modules `05` and `06`)

You can navigate to any of these and modify the code within the self-contained **R Markdown** code chunks. Entire **R Markdown** files can be knit to HTML, where a separate window will show the rendered document.

You can navigate to any of these and modify the code within the self-contained **R Markdown** code chunks. Entire **R Markdown** files can be knit to HTML, where a separate window will show the rendered document.

### The **pointblank** Installation

Normally you would install **pointblank** on your system by using `install.packages()`:

```{r eval=FALSE}
# install.packages("pointblank")
```

For this workshop, however, we are going to use the development version of **pointblank** and install it from GitHub with `devtools::install_github()`.

```{r eval=FALSE}
# devtools::install_github("rich-iannone/pointblank")
```



