---
title: "Scaling up Data Validation with the Multiagent"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pointblank)
library(tidyverse)
```

## Intro

If your data quality process involves more data validation runs and you have a large variety of tables to validate, we can take advantage of functions available in **pointblank** to make all that possible.

### Writing agents to disk

The `x_write_disk()` function lets you write a **pointblank** agent to disk. This is very useful if you want to continually save the interrogation results as part of a larger data quality process. Here's an example where the agent is saved to disk with the date as part of the filename.

```{r eval=FALSE}

# Create the agent, develop a validation plan, interrogate
agent <-
  create_agent(
    tbl = ~ small_table,
    tbl_name = "small_table",
    label = "Daily check of `small_table`.",
    actions = action_levels(
      warn_at = 0.10,
      stop_at = 0.25,
      notify_at = 0.35
    )
  ) |>
  col_exists(columns = vars(date, date_time)) |>
  col_vals_regex(
    columns = b,
    regex = "[0-9]-[a-z]{3}-[0-9]{3}"
  ) |>
  rows_distinct() |>
  col_vals_gt(columns = d, value = 100) |>
  col_vals_lte(columns = c, value = 5) |>
  interrogate()

# Save the agent to disk with `x_write_disk()`; append the date
x_write_disk(
  agent,
  filename = affix_date(filename = "agent-small_table"),
  path = "small_table_tests"
)
```

### Reading agents from disk

We have this on disk as `"small_table_tests/agent-small_table_2022-10-13"`. We can read this from disk using the `x_read_disk()` function (it recreates the object).

```{r}
agent_2022_10_13 <- 
  x_read_disk(
    filename = "agent-small_table_2022-10-13",
    path = "small_table_tests"
  )
```

We can get the data validation report from it.

```{r}
agent_2022_10_13
```

### Creating a 'multiagent' to get a combined data validation report

A common task might be to see how data quality is changing over time. If you have multiple saved agents that check the same table, we can make a combined validation report that shows all of those validations.

We actually have five saved agents in the `"small_table_tests"` directory:

- `"agent-small_table_2022-10-13"`
- `"agent-small_table_2022-10-14"`
- `"agent-small_table_2022-10-15"`
- `"agent-small_table_2022-10-16"`
- `"agent-small_table_2022-10-17"`

Let's get them all into a single report. We do this by generating a `multiagent` and that object has it's own `get_multiagent_report()` function for customizing the layout and content of the report.

```{r}
multiagent <- 
  create_multiagent(
    x_read_disk("small_table_tests/agent-small_table_2022-10-13"),
    x_read_disk("small_table_tests/agent-small_table_2022-10-14"),
    x_read_disk("small_table_tests/agent-small_table_2022-10-15"),
    x_read_disk("small_table_tests/agent-small_table_2022-10-16"),
    x_read_disk("small_table_tests/agent-small_table_2022-10-17")
  )
```

We can get a combined data validation report from it. By default, all validation reports are stacked together in the `"long"` display mode.

```{r}
multiagent
```

With `get_multiagent_report()` we can customize the reporting. Here, we will choose the `"wide"` display mode and provide a custom title.

```{r}
get_multiagent_report(
  multiagent,
  display_mode = "wide",
  title = "Wide report from **Multiple** Table Validations"
)
```

------

### SUMMARY

1. We can save agents to disk and read them back. This is good for keeping records of data quality and all data/reporting is preserved.
2. Multiple agents can be combined together, generating specialized reports that can show the validations of multiple table (long display) of the evolution of data quality for a single table (wide display).
